Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-11 paper2.qmd
section:theoretical; process:write,figures

<!-- 2. the preliminary SCM for CJ -->
@fig-cj04 illustrates the preliminary integration of theoretical principles into the conceptual-population model. Specifically, @fig-cj04_scm shows the SCM that underlies the motivating example. This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{abhijk})$ to several other variables. In this SCM, the subscripts $\{a,b\}$ represent the compared texts (stimuli), with $a \neq b$. The indices $\{h,i\}$ label the students who wrote the texts (individuals), where $h = i$ is possible. Moreover, the index $j$ denotes the judge, and $k$ denotes the judgment number, which accounts for experimental cases where judges perform multiple judgments of the same texts written by the same students, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376].

The other variables in [SCM -@fig-cj04_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory). Additionally, based on the evidence presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, we incorporate the judges' biases $(B_{kj})$ into the model. Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$. Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

<!-- 3. What are the innovations of this? -->
By incorporating judges' biases into the conceptual-population model, we assume that additional relevant characteristics of the stimuli are unlikely to be equally weighted or opposing. Thus, these characteristics may exert an uneven influence on judges' perceptions, creating biases in their judgments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp. 346], hence the need for their integration. The full statistical elaboration of this point is presented in @sec-statistical.

@@

::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
\end{aligned}
$$

@@

<!-- 4. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables. The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020]. Thus, this study uses the magnified representation.

<!-- 5. the preliminary DAG for CJ -->
@fig-cj04_dag displays the DAG associated with [SCM -@fig-cj03_scm]. The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ experiments. It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the the text's discriminal processes, their discriminal difference, and the judges' biases.

<!-- 6. The problem of continuing this way -->
It is easy to see that representing the CJ system using indices for a single pair of individuals and stimuli, along with the indices for the judgment number and the judge responsible for the comparison, can quickly become intractable and cumbersome. Thus, to maintain generality and simplify the notation, certain indices are grouped into vectors or matrices. First, we define the matrices $IA$ and $JK$ as:

$$
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
$$

Each row of $IA$ represents a unique pairing of an individual $\{h,i\}$ with a stimulus $\{a,b\}$. Here, $n_{I}$ represents the number of individuals and $n_{A}$ the number of stimulus within individuals in the population. Similarly, each row of $JK$ represents a unique pairing of a judge $j$ with a judgment $k$. Here, $n_{J}$ represents the number of judges and $n_{K}$ the number of judgments performed by each judge in the population.

<!-- Then, the study defines the matrices of stimuli and individual comparisons, $P \in \mathbb{N}^{m \times 2}$ and $R \in \mathbb{N}^{n \times 2}$, respectively. Here, each row of $P$ represents a pairing $\{a,b\}$ of stimuli indices and each row of $R$ represents a pairing $\{h,i\}$ of individual indices, i.e., -->

<!-- $$ -->
<!-- P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\} -->
<!-- $$ -->

<!-- Here, $m$ and $n$ represents the number of unique pairwise comparisons between stimuli, and unique pairwise comparison per individuals, which corresponds to the number of ways to choose two elements from $U_{A}$ and $U_{I}$, typically given by: -->

<!-- $$ -->
<!-- m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!} -->
<!-- $$ -->

<!-- Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in   -->

@@

## Statistics and Causal inference {#sec-appendixB .appendix appendix-style=plain}

>>>>

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] illustrates the preliminary integration of theoretical principles into the conceptual-population model. This SCM maps the relationship between the conceptual-population outcome, $(O^{cp}_{abhijk})$, and several related variables. The subscripts $a$ and $b$ represent the compared texts (stimuli), while the indices $h$ and $i$ label the students who authored them (individuals). Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges compare the same texts written by the same students multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. It is important to note that the indexing system allows for comparisons between different texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student’s text to itself ($h = i$, $a = b$), as such a comparison lacks practical meaning.

The other variables in [SCM -@fig-cj03_scm] include the texts' discriminal processes $(T_{ahjk}, T_{bijk})$ and their discriminal difference $(D_{abhijk})$, in line with Thurstone's theory (see @sec-thurstone_theory). Additionally, based on the arguments presented in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}$. Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
$$

@@

<!-- 4. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another showing the direction of causal influence. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. The *standard representation* of DAGs typically omits exogenous variables for simplicity, while the *magnified representation* includes them to highlight potential issues related to conditioning and confounding [@Cinelli_et_al_2020]. This section adopts the magnified representation to enhance clarity.

<!-- 5. the preliminary DAG for CJ -->
@fig-cj03_dag displays the DAG associated with [SCM -@fig-cj03_scm]. The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory and CJ's experiments. It also distinguishes between observable endogenous variables, such as the outcome, and variables considered latent according to the theory and practice of CJ, such as the text's discriminal processes, their discriminal difference, and the judges' biases.

<!-- 6. The problem of continuing this way -->
Although SCMs and DAGs do not necessitate specifying the data structure of the modeled system, defining these structures can sometimes enhance their clarity and simplify their notation. Thus, we organize certain indices into vectors or matrices. First, we define the matrices $IA$ and $JK$ as follows:

$$
IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N} \; ; \quad
JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
$$ {#eq-mat12}

where each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$. Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual. Thus, the matrix $IA$ is composed of natural numbers $\mathbb{N}$, with $n_{I} \cdot n_{A}$ rows and $2$ columns. Similarly, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment $k$. In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs. Therefore, the matrix $JK$ is also composed of natural numbers $\mathbb{N},$ with $n_{J} \cdot n_{K}$ rows and $2$ columns.

Importantly, since the discriminal difference of the stimuli becomes an observable outcome only through judges' perceptions, it is necessary to define the matrix $RPUV$, where each row of the matrix uniquely pairs the corresponding rows from the matrices $IA$ and $JK$ as follows:

$$
RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & a & i & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
$$ {#eq-mat3}

Here, the $RPUV$ matrix is also composed of natural numbers $\mathbb{N}$, and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. The dimension of the matrix results from the conceptual-population model's assumption that researchers have access to data representing all repeated comparisons made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Here, the term ${n_{I} \cdot n_{a} \choose 2}$ represents the Binomial coefficient, which quantifies the number of individual-stimuli comparison pairs, and it is defined as:

$$
{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
$$

Using the motivating example outlined in @sec-theoretical, we can visualize the shape of these structures. For illustrative purposes, assume the conceptual population consists of $n_{I} = 5$ individuals, each with $n_{A} = 2$ available stimuli. Additionally, assume that each pair of stimuli is evaluated by $n_{J} = 3$ judges, with each judge making $n_{K} = 3$ comparative judgments. Given this setup, the matrices in ([-@eq-mat12]) and ([-@eq-mat3]) are represented as follows:

$$
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix}\; ; \quad
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} \; ; \quad
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
$$

With these matrices we can express an equivalent version of [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a vectorized form.

@@

## Statistical and Causal inference {#sec-appendixB .appendix appendix-style=plain}
